---
title: "AGU_GEDI_Workshop"
author: "Melissa Rose, Ivan Gonzalez, Jenna Keanny"
date: '2022-11-21'
output: html_document
---

SECTION 0: PREPARE R ENVIRONMENT
0.1: Libraries
```{r, warning=FALSE, message=FALSE}
#define which libraries to use
library(raster)   
library(devtools) 
library(sf) 
library(sp)
library(data.table)
library(lidR)
library(httr)
library(tidyr)
library(curl)
library(hdf5r)
library(leaflet)
library(leafsync)
library(ggplot2)
library(RColorBrewer)
library(rasterVis)
library(viridis)
```

0.2: Directory and Subdirectories 
```{r}
#define folder paths
wd <- getwd()

gL1B_dir <- './data/GEDI01_B.002/'
gL2A_dir <- './data/GEDI02_A.002/'
gL2B_dir <- './data/GEDI02_B.002/'
```

0.3: Source Scripts
```{r, warning=FALSE, message=FALSE}
#source rGEDI functions in scripts directory
source('./scripts/gedi_functions.R')
```


SECTION 1: DOWNLOAD GEDI DATA FOR A REGION OF INTEREST
1.1: Define region of interest and date range 
```{r}
#ROI: Columbia
#specify bounding box coordinates for ROI
ll_lon <- -75  #lower left longitude 
ll_lat <- 0    #lower left latitude 
ur_lon <- -74  #upper right longitude
ur_lat <- 1    #upper right latitude 

#convert into bounding box coordinates
bbox <- paste(ll_lon, ll_lat, ur_lon, ur_lat, sep = ',')

#specify the date range to filter orbits by
daterange=c("2020-05-01","2020-07-31")
```

1.2: Visualize region of interest
```{r}
#view footprint locations for entire orbit on map 
leaflet() %>%
  addRectangles(
    lng1=ll_lon, lat1=ll_lat,
    lng2=ur_lon, lat2=ur_lat,
    color = "grey",
    fillColor = "transparent") %>%
  addScaleBar(options = list(imperial = FALSE)) %>%
  addProviderTiles(providers$Esri.WorldImagery) %>%
  addLegend(colors = "grey", labels = "Region of Interest") 
```

1.3: Use gedi_finder to find overlapping GEDI orbits 
```{r, warning=FALSE, message=FALSE}
#create a list of all overlapping orbits for each GEDI data product 
gL1B <- gedi_finder('GEDI01_B.002', bbox)  
gL2A <- gedi_finder('GEDI02_A.002', bbox)
gL2B <- gedi_finder('GEDI02_B.002', bbox)
```

1.4 Use gedi_temp_filter to subset orbit list within date range
```{r}
#subset orbit list using daterange
gL1B <- gedi_temp_filter(gL1B, 'GEDI01_B.002', daterange)
gL2A <- gedi_temp_filter(gL2A, 'GEDI02_A.002', daterange)
gL2B <- gedi_temp_filter(gL2B, 'GEDI02_B.002', daterange)
```

1.5: Use gedi_download to download GEDI data (DO NOT RUN)
```{r, eval=FALSE}
#download data product to their output data directory 
gedi_download(filepath=gL1B,outdir=gL1B_dir)
gedi_download(filepath=gL2A,outdir=gL2A_dir)
gedi_download(filepath=gL2B,outdir=gL2B_dir)
```


SECTION 2: CLIP GEDI DATA PRODUCTS TO REGION OF INTEREST
2.1 Read GEDI data (DO NOT RUN)
```{r, eval=FALSE}
#read the level 1B, 2A, and 2B for a single orbit
gedilevel1B <- readLevel1B(level1Bpath = paste0(gL1B_dir,"GEDI01_B_2020160005619_O08428_03_T01337_02_005_01_V002.h5"))
gedilevel2A <- readLevel2A(level2Apath = paste0(gL2A_dir,"GEDI02_A_2020160005619_O08428_03_T01337_02_003_01_V002.h5"))
gedilevel2B <- readLevel2B(level2Bpath = paste0(gL2B_dir,"GEDI02_B_2020160005619_O08428_03_T01337_02_003_01_V002.h5"))
```

2.2 Clip data within bbox coordinates (DO NOT RUN)
```{r, eval=FALSE}
gedilevel1B <- clipLevel1B(gedilevel1B, ll_lon, ur_lon, ll_lat, ur_lat,
                                output = paste0(gL1B_dir,"GEDI01_B_2020160005619_O08428_03_T01337_02_005_01_V002_clip.h5"))
gedilevel2A <- clipLevel2A(gedilevel2A, ll_lon, ur_lon, ll_lat, ur_lat, 
                                output = paste0(gL2A_dir,"GEDI02_A_2020160005619_O08428_03_T01337_02_003_01_V002_clip.h5"))
gedilevel2B <- clipLevel2B(gedilevel2B, ll_lon, ur_lon, ll_lat, ur_lat, 
                                output = paste0(gL2B_dir,"GEDI02_B_2020160005619_O08428_03_T01337_02_003_01_V002_clip.h5"))
```

###START HERE###
```{r}
gedilevel1B <- readLevel1B(level1Bpath = paste0(gL1B_dir,"GEDI01_B_2020160005619_O08428_03_T01337_02_005_01_V002_clip.h5"))
gedilevel2A <- readLevel2A(level2Apath = paste0(gL2A_dir,"GEDI02_A_2020160005619_O08428_03_T01337_02_003_01_V002_clip.h5"))
gedilevel2B <- readLevel2B(level2Bpath = paste0(gL2B_dir,"GEDI02_B_2020160005619_O08428_03_T01337_02_003_01_V002_clip.h5"))
```

2.3 Use GEDI L2A quality flag to select only quality shots
```{r}
#read GEDI L2A data into data.table
level2AM <- getLevel2AM(gedilevel2A)
dim(level2AM)

#convert shot_number from integer to character
level2AM$shot_number <- paste0(level2AM$shot_number)

#filtering for quality_flag == 1
level2AM_qual <- level2AM[quality_flag == 1]
dim(level2AM_qual)
```

2.4 Visualize location of quality filtered GEDI footprints 
```{r}
#view footprint locations for entire orbit on map 
leaflet() %>%
  addRectangles(
    lng1=ll_lon, lat1=ll_lat,
    lng2=ur_lon, lat2=ur_lat,
    color = "grey",
    fillColor = "transparent") %>%
  addCircleMarkers(level2AM[quality_flag == 0, lon_lowestmode],
                   level2AM[quality_flag == 0, lat_lowestmode],
                   radius = 1,
                   opacity = 0.5,
                   color = "red", 
                   popup = paste0("shot_number: ", level2AM$shot_number))  %>%
  addCircleMarkers(level2AM_qual$lon_lowestmode,
                   level2AM_qual$lat_lowestmode,
                   radius = 1,
                   opacity = 0.5,
                   color = "blue", 
                   popup = paste0("shot_number: ", level2AM_qual$shot_number))  %>%
  addScaleBar(options = list(imperial = FALSE)) %>%
  addProviderTiles(providers$Esri.WorldImagery) %>%
  addLegend(colors = c("grey", "blue", "red"), labels = c("Region of Interest", "Quality GEDI Footprints", "Non-Quality GEDI Footprints"))
```

2.4 Select individual footprints along a structurally heterogeneous gradient
```{r}
sample_region <- level2AM_qual[shot_number %in% c("84280500300242755", "84280500300242756", "84280500300242757", "84280500300242758", "84280500300242759")]

#view footprint in sample region 
leaflet() %>%
  addRectangles(
    lng1=min(sample_region$lon_lowestmode)-0.00025, lat1=min(sample_region$lat_lowestmode)-0.00025,
    lng2=max(sample_region$lon_lowestmode)+0.00025, lat2=max(sample_region$lat_lowestmode)+0.00025,
    color = "grey",
    fillColor = "transparent") %>%
  addCircleMarkers(sample_region$lon_lowestmode,
                   sample_region$lat_lowestmode,
                   radius = 25,
                   weight = 2,
                   fillColor = "transparent",
                   color = "blue", 
                   popup = paste0("shot_number: ", sample_region$shot_number))  %>%
  addScaleBar(options = list(imperial = FALSE)) %>%
  addProviderTiles(providers$Esri.WorldImagery) %>%
  addLegend(colors = c("grey", "blue"), labels = c("Sample Region", "GEDI Footprints")) 
```

SECTION 3: Visualize GEDI Data Products 
3.1 Plot GEDIL1B Full Waveform 
```{r}
#select shot_numbers for different cover types
shotnum_forest <- "84280500300242756"
shotnum_edge <- "84280500300242757"
shotnum_grass <- "84280500300242758"

#get waveform data for each selected footprint
wf_forest <- getLevel1BWF(gedilevel1B, shot_number=shotnum_forest)
wf_edge <- getLevel1BWF(gedilevel1B, shot_number=shotnum_edge)
wf_grass <- getLevel1BWF(gedilevel1B, shot_number=shotnum_grass)
```

```{r}
#plot raw waveforms
par(mfrow = c(1,1), mar=c(4,4,1,1), cex.axis = 1.5)

plot(wf_forest, relative=FALSE, polygon=TRUE, type="l", lwd=2, col="blue",
    xlab="Waveform Amplitude", ylab="Elevation (m)", main="Raw Waveform: Forest", xlim=c(200,500), ylim=c(200,260))

plot(wf_edge, relative=FALSE, polygon=TRUE, type="l", lwd=2, col="blue",
    xlab="Waveform Amplitude", ylab="Elevation (m)", main="Raw Waveform: Edge", xlim=c(200,500), ylim=c(200,260), add = TRUE)

plot(wf_grass, relative=FALSE, polygon=TRUE, type="l", lwd=2, col="blue",
    xlab="Waveform Amplitude", ylab="Elevation (m)", main="Raw Waveform: Grass", xlim=c(200,500), ylim=c(200,260), add = TRUE)
```

3.2 Plot GEDIL2A Elevation and Height Metrics 
```{r}
#plot waveform with GEDIL2A relative height metrics
plotWFMetrics(gedilevel1B, gedilevel2A, shotnum_forest, rh=c(25, 50, 75, 90))
plotWFMetrics(gedilevel1B, gedilevel2A, shotnum_edge, rh=c(25, 50, 75, 90))
plotWFMetrics(gedilevel1B, gedilevel2A, shotnum_grass, rh=c(25, 50, 75, 90))
```

3.3 Plot GEDIL2B Plant Area Index (PAI) and Plant Area Volume Density (PAVD)
```{r}
#get canopy cover and vertical profile metrics
level2BVPM <- getLevel2BVPM(gedilevel2B)
dim(level2BVPM)

#filtering for quality_flag == 1
level2BVPM <- level2BVPM[l2b_quality_flag == 1]
dim(level2BVPM)

#removing NA values 
level2BVPM$pai[level2BVPM$pai==-9999]<-NA # assigning NA to -9999
dim(level2BVPM)
```


```{r}
#get pai profile metrics 
level2BPAIProfile <- getLevel2BPAIProfile(gedilevel2B)
dim(level2BPAIProfile)

#filtering for quality_flag == 1
level2BPAIProfile <- level2BPAIProfile[l2b_quality_flag == 1]
dim(level2BPAIProfile)

#convert shot_number from integer to character
level2BPAIProfile$shot_number <- paste0(level2BPAIProfile$shot_number)

#plot Level2B PAI Profile
#gPAIprofile <- plotPAIProfile(level2BPAIProfile, beam="BEAM0101", elev=TRUE)
```

```{r}
#get pavd profile metrics 
level2BPAVDProfile <- getLevel2BPAVDProfile(gedilevel2B)
dim(level2BPAVDProfile)

#filtering for quality_flag == 1
level2BPAVDProfile <- level2BPAVDProfile[l2b_quality_flag == 1]
dim(level2BPAVDProfile)

#plot Level2B PAVD Profile
#gPAVDprofile<-plotPAVDProfile(level2BPAVDProfile, beam="BEAM0101", elev=TRUE)
```

SECTION 4: Create Raster Layers 
4.1 Compute Descriptive Statistics 
```{r}
#compute a series of statistics for GEDI relative height metrics 
rh100metrics_st <- polyStatsLevel2AM(level2AM_qual, func=SetOfMetrics(rh100))

#compute a series of statistics for GEDI canopy cover metrics 
cover_metrics_st <- polyStatsLevel2BVPM(level2BVPM, func=SetOfMetrics(cover))
```

4.1 Compute Descriptive Height Stats 
```{r}
#compute a series of statistics for GEDI relative height metrics 
rh100metrics <- gridStatsLevel2AM(level2AM_qual, func=SetOfMetrics(rh100), res=0.010)
```

```{r}
#plot rh100 stats
rh100maps <- levelplot(rh100metrics,
                     layout=c(2, 2),
                     margin=FALSE,
                     xlab = "Longitude (degree)", ylab = "Latitude (degree)",
                     colorkey=list(
                       space='right',
                       labels=list(at=seq(0, 46, 2), font=4),
                       axis.line=list(col='black'),
                       width=1),
                     par.settings=list(
                       strip.border=list(col='gray'),
                       strip.background=list(col='gray'),
                       axis.line=list(col='gray')
                     ),
                     scales=list(draw=TRUE),
                     col.regions=viridis,
                     at=seq(0, 46, len=101),
                     names.attr=c("rh100 min","rh100 max","rh100 mean", "rh100 sd"))
rh100maps
```

4.2 Compute Descriptive Canopy Cover Stats 
```{r}
#compute a series of statistics for GEDI plant area index
pai_metrics <- gridStatsLevel2BVPM(level2BVPM = level2BVPM, func=SetOfMetrics(pai), res=0.010)
fhd_metrics <- gridStatsLevel2BVPM(level2BVPM = level2BVPM, func=SetOfMetrics(fhd_normal), res=0.010)
cover_metrics <- gridStatsLevel2BVPM(level2BVPM = level2BVPM, func=SetOfMetrics(cover), res=0.010)
```

```{r}
#plot pai stats
pai_maps <- levelplot(pai_metrics,
                    layout=c(2, 2),
                    margin=FALSE,
                    xlab = "Longitude (degree)", ylab = "Latitude (degree)",
                    colorkey=list(
                      space='right',
                      labels=list(at=seq(0, 8, 0.2), font=4),
                      axis.line=list(col='black'),
                      width=1),
                    par.settings=list(
                      strip.border=list(col='gray'),
                      strip.background=list(col='gray'),
                      axis.line=list(col='gray')
                    ),
                    scales=list(draw=TRUE),
                    col.regions=viridis,
                    at=seq(0, 8, len=101),
                    names.attr=c("PAI min","PAI max","PAI mean", "PAI sd"))
pai_maps
```

```{r}
#plot fhd stats
fhd_maps <- levelplot(fhd_metrics,
                    layout=c(2, 2),
                    margin=FALSE,
                    xlab = "Longitude (degree)", ylab = "Latitude (degree)",
                    colorkey=list(
                      space='right',
                      labels=list(at=seq(0, 3.5, 0.25), font=4),
                      axis.line=list(col='black'),
                      width=1),
                    par.settings=list(
                      strip.border=list(col='gray'),
                      strip.background=list(col='gray'),
                      axis.line=list(col='gray')
                    ),
                    scales=list(draw=TRUE),
                    col.regions=viridis,
                    at=seq(0, 3.5, len=101),
                    names.attr=c("FHD min","FHD max","FHD mean", "FHD sd"))
fhd_maps
```


```{r}
#plot cover stats
cover_maps <- levelplot(cover_metrics,
                    layout=c(2, 2),
                    margin=FALSE,
                    xlab = "Longitude (degree)", ylab = "Latitude (degree)",
                    colorkey=list(
                      space='right',
                      labels=list(at=seq(0, 1, 0.05), font=4),
                      axis.line=list(col='black'),
                      width=1),
                    par.settings=list(
                      strip.border=list(col='gray'),
                      strip.background=list(col='gray'),
                      axis.line=list(col='gray')
                    ),
                    scales=list(draw=TRUE),
                    col.regions=viridis,
                    at=seq(0, 1, len=101),
                    names.attr=c("CCF min","CCF max","CCF mean", "CCF sd"))
cover_maps
```