---
title: "AGU_GEDI_Workshop"
author: "Melissa Rose, Ivan Gonzalez, Jenna Keanny"
date: '2022-11-21'
output: html_document
header-includes:
  \usepackage{fvextra}
  \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
---



## SECTION 0: PREPARE R ENVIRONMENT
0.1: Libraries

(Do not need to run this chunk when running in Binder)
```{r, warning=FALSE, message=FALSE, eval=F}

#Install libraries if not already installed

libs <- c('raster', 'devtools', 'sf', 'sp', 'data.table', 'dplyr', 'lidR', 
          'httr', 'tidyr', 'curl', 'hdf5r', 'leaflet', 'leaflet.extras', 
          'leafsync', 'ggplot2', 'rgeos', 'RColorBrewer', 'rasterVis',
          'viridis', 'rdgal', 'rhdf5', 'dismo', 'sdm')

sapply(libs, FUN = function(x){
  if (!x %in% installed.packages()){
    install.packages(x)
  } 
})

#it is a good practice to call the installAll function (just one time when the sdm package is installed), that tries to install all the packages that may be needed somewhere in the sdm package.
sdm::installAll()
```

```{r load libraries, warning=FALSE, message=FALSE}
#Define which libraries to use
library(raster)       ## spatial tools    
library(devtools)     ## tools for building and installing libraries
library(sf)           ## spatial tools 
library(sp)           ## spatial tools
library(data.table)   ## data / table manipulation
library(lidR)         ## lidar manipulation tools
library(httr)         ## methods for web files interaction 
library(tidyr)        ## data / table manipulation
library(curl)         ## download files from the web
library(hdf5r)        ## reading hdf5 files
library(leaflet)      ## color palettes
library(leafsync)     ## interactive maps
library(ggplot2)      ## nice plot tools
library(rgeos)        ## geometric for vector data
library(RColorBrewer) ## color palettes
library(rasterVis)    ## raster viz tools
library(viridis)      ## color palettes
library(rgdal)        ## readOGR: Read OGR vector maps into Spatial objects
library(rhdf5)        ## R Interface to HDF5
library(dismo)        ## Species Distribution Modeling
library(sdm)          ## Species Distribution Modelling
library(randomForest) ## For sdm methd 'rf'
#library(rGEDI)       ## GEDI functions in R

```

0.2: Directory and Subdirectories 
```{r definw working directory}
#define folder paths
wd <- getwd()

gL1B_dir <- './data/GEDI01_B.002/'
gL2A_dir <- './data/GEDI02_A.002/'
gL2B_dir <- './data/GEDI02_B.002/'
```

0.3: Source Scripts
```{r source functions, warning=FALSE, message=FALSE}
#source rGEDI functions in scripts directory
source('./scripts/gedi_functions.R')
```


## SECTION 1: DOWNLOAD GEDI DATA FOR A REGION OF INTEREST
1.1: Define region of interest and date range 
```{r bbox}
#ROI: Colombia
#specify bounding box coordinates for ROI
ll_lon <- -75  #lower left longitude 
ll_lat <- 0    #lower left latitude 
ur_lon <- -74  #upper right longitude
ur_lat <- 1    #upper right latitude 

#convert into bounding box coordinates
bbox <- paste(ll_lon, ll_lat, ur_lon, ur_lat, sep = ',')

#specify the date range to filter orbits by
daterange <- c("2020-05-01","2020-07-31")
```

1.2: Visualize region of interest
```{r viz study area}
#view footprint locations for entire orbit on map 
leaflet() %>%
  addRectangles(
    lng1=ll_lon, lat1=ll_lat,
    lng2=ur_lon, lat2=ur_lat,
    color = "grey",
    fillColor = "transparent") %>%
  addScaleBar(options = list(imperial = FALSE)) %>%
  addProviderTiles(providers$Esri.WorldImagery) %>%
  addLegend(colors = "grey", labels = "Region of Interest") 
```

1.3: Use gedi_finder to find overlapping GEDI orbits 
```{r find orbits, warning=FALSE, message=FALSE}
#create a list of all overlapping orbits for each GEDI data product 
gL1B <- gedi_finder('GEDI01_B.002', bbox)  
gL2A <- gedi_finder('GEDI02_A.002', bbox)
gL2B <- gedi_finder('GEDI02_B.002', bbox)
```

1.4 Use gedi_temp_filter to subset orbit list within date range
```{r filter orbits}
#subset orbit list using daterange
gL1B <- gedi_temp_filter(gL1B, 'GEDI01_B.002', daterange)
gL2A <- gedi_temp_filter(gL2A, 'GEDI02_A.002', daterange)
gL2B <- gedi_temp_filter(gL2B, 'GEDI02_B.002', daterange)
```

1.5: Use gedi_download to download GEDI data (DO NOT RUN -- takes a while and we have some files downloades already)
```{r download, eval=FALSE}
#download data product to their output data directory 
#This functions will download heavy files (>1GB each) that can take long time

# gedi_download(filepath=gL1B,outdir=gL1B_dir)
# gedi_download(filepath=gL2A,outdir=gL2A_dir)
# gedi_download(filepath=gL2B,outdir=gL2B_dir)
```


SECTION 2: CLIP GEDI DATA PRODUCTS TO REGION OF INTEREST
Stop here -------------------------------
2.1 Read GEDI data (DO NOT RUN)
```{r load original files, eval=FALSE}
#read the level 1B, 2A, and 2B for a single orbit
## Remove the "#" symbol in the next lines in order to run them
# gedilevel1B <- readLevel1B(level1Bpath = paste0(gL1B_dir,"GEDI01_B_2020160005619_O08428_03_T01337_02_005_01_V002.h5"))
# gedilevel2A <- readLevel2A(level2Apath = paste0(gL2A_dir,"GEDI02_A_2020160005619_O08428_03_T01337_02_003_01_V002.h5"))
# gedilevel2B <- readLevel2B(level2Bpath = paste0(gL2B_dir,"GEDI02_B_2020160005619_O08428_03_T01337_02_003_01_V002.h5"))
```

2.2 Clip data within bbox coordinates (DO NOT RUN)
```{r filter h5 files , eval=FALSE}
## Remove the "#" symbol in the next lines in order to run them
# gedilevel1B <- clipLevel1B(gedilevel1B, ll_lon, ur_lon, ll_lat, ur_lat, output = paste0(gL1B_dir,"GEDI01_B_2020160005619_O08428_03_T01337_02_005_01_V002_clip.h5"))
# gedilevel2A <- clipLevel2A(gedilevel2A, ll_lon, ur_lon, ll_lat, ur_lat, output = paste0(gL2A_dir,"GEDI02_A_2020160005619_O08428_03_T01337_02_003_01_V002_clip.h5"))
# gedilevel2B <- clipLevel2B(gedilevel2B, ll_lon, ur_lon, ll_lat, ur_lat, output = paste0(gL2B_dir,"GEDI02_B_2020160005619_O08428_03_T01337_02_003_01_V002_clip.h5"))
```
Restart here -------------------------------

Read clipped files into R 
```{r import GEDI data}
gedilevel1B <- readLevel1B(level1Bpath = paste0(gL1B_dir,"GEDI01_B_2020160005619_O08428_03_T01337_02_005_01_V002_clip.h5"))
gedilevel2A <- readLevel2A(level2Apath = paste0(gL2A_dir,"GEDI02_A_2020160005619_O08428_03_T01337_02_003_01_V002_clip.h5"))
gedilevel2B <- readLevel2B(level2Bpath = paste0(gL2B_dir,"GEDI02_B_2020160005619_O08428_03_T01337_02_003_01_V002_clip.h5"))
```

2.3 Use GEDI L2A quality flag to select only quality shots
```{r manipulate L2A}
#read GEDI L2A data into data.table
level2AM <- getLevel2AM(gedilevel2A)
head(level2AM, 3) ## Check the data structure
dim(level2AM)

#convert shot_number from integer to character
level2AM$shot_number <- paste0(level2AM$shot_number)

#filtering for quality_flag == 1
level2AM_qual <- level2AM[quality_flag == 1]
dim(level2AM_qual)
```

2.4 Visualize location of quality filtered GEDI footprints 
```{r plot original points}
#view footprint locations for entire orbit on map 
leaflet() %>%
  addRectangles(
    lng1=ll_lon, lat1=ll_lat,
    lng2=ur_lon, lat2=ur_lat,
    color = "grey",
    fillColor = "transparent") %>%
  addCircleMarkers(level2AM[quality_flag == 0, lon_lowestmode],
                   level2AM[quality_flag == 0, lat_lowestmode],
                   radius = 1,
                   opacity = 0.5,
                   color = "red", 
                   popup = paste0("shot_number: ", level2AM$shot_number))  %>%
  addCircleMarkers(level2AM_qual$lon_lowestmode,
                   level2AM_qual$lat_lowestmode,
                   radius = 1,
                   opacity = 0.5,
                   color = "blue", 
                   popup = paste0("shot_number: ", level2AM_qual$shot_number))  %>%
  setView(median(level2AM_qual$lon_lowestmode),
          median(level2AM_qual$lat_lowestmode), 
          zoom = 10) %>%
  addScaleBar(options = list(imperial = FALSE)) %>%
  addProviderTiles(providers$Esri.WorldImagery) %>%
  addLegend(colors = c("grey", "blue", "red"), labels = c("Region of Interest", "Quality GEDI Footprints", "Non-Quality GEDI Footprints"))
```


```{r explore beams}
# View beams based on their type
# BEAM0000,BEAM0001,BEAM0010,BEAM0011 are Coverage Beams. 
# BEAM0101,BEAM0110,BEAM1000,BEAM1011 are Full Power Beams
level2AM_qual$beamType <- ifelse(level2AM_qual$beam %in% c('BEAM0000','BEAM0001','BEAM0010','BEAM0011'), 'Coverage', 'Power')

#view footprint locations for entire orbit on map 
leaflet() %>%
  addCircleMarkers(level2AM_qual$lon_lowestmode,
                   level2AM_qual$lat_lowestmode,
                   radius = 1,
                   opacity = 0.5,
                   color = c('red', 'blue')[as.numeric(as.factor(level2AM_qual$beamType))], 
                   popup = paste0(level2AM_qual$beamType, ' beam - ', level2AM_qual$beam, 
                                  '<br>', level2AM_qual$shot_number ))  %>%

  addScaleBar(options = list(imperial = FALSE)) %>%
  addProviderTiles(providers$Esri.WorldImagery) %>%
  addLegend(colors = c( "blue", "red"), 
            labels = c("Power beam", "Coverage beam"))

```

```{r plot beams shots}
# Plot power and coverage shots ID from previous interactive map
plotVertProfiles <- plot_rh_profiles(level2AM_qual[shot_number %in%
                                                     c("84280500300243078", "84280300300243183")], # power and coverage
                                     idcolumn = 'beamType', col_vector = c('blue', 'red'))
```

2.4 Select individual footprints along a structurally heterogeneous gradient
```{r veg gradient}
sample_shots <- c("84280500300242755", "84280500300242756", "84280500300242757", "84280500300242758", "84280500300242759")
sample_region <- level2AM_qual[shot_number %in% sample_shots]
sample_region$sortid <- 1:nrow(sample_region)

#view footprint in sample region 
colorRamp <- colorNumeric( palette = "viridis", domain = sample_region$sortid, reverse = TRUE)
pointColors <- colorRamp(sample_region$sortid)

leaflet() %>%
  addRectangles(
    lng1=min(sample_region$lon_lowestmode)-0.00025, lat1=min(sample_region$lat_lowestmode)-0.00025,
    lng2=max(sample_region$lon_lowestmode)+0.00025, lat2=max(sample_region$lat_lowestmode)+0.00025,
    color = "grey",
    fillColor = "transparent") %>%
  addCircleMarkers(sample_region$lon_lowestmode,
                   sample_region$lat_lowestmode,
                   radius = 25,
                   weight = 5,
                   fillColor = "transparent",
                   color = pointColors, 
                   popup = paste0("shot_number: ", sample_region$shot_number))  %>%
  addScaleBar(options = list(imperial = FALSE)) %>%
  addProviderTiles(providers$Esri.WorldImagery) %>%
  addLegend(colors = c("grey", "blue"), labels = c("Sample Region", "GEDI Footprints")) 
```

Now lets see their vertical profiles
```{r plot vertical profiles}
plotVertProfiles <- plot_rh_profiles(sample_region, idcolumn = 'shot_number', col_vector = pointColors)
```

## SECTION 3: Visualize GEDI Data Products 
3.1 Plot GEDIL1B Full Waveform 
```{r get waveforms L1B}
#select shot_numbers for different cover types
shotnum_forest <- "84280500300242755"
shotnum_edge <- "84280500300242757"
shotnum_grass <- "84280500300242759"
sample_shotnums <- c(shotnum_forest, shotnum_edge, shotnum_grass)

#get waveform data for each selected footprint
wf_forest <- getLevel1BWF(gedilevel1B, shot_number=shotnum_forest)
wf_edge <- getLevel1BWF(gedilevel1B, shot_number=shotnum_edge)
wf_grass <- getLevel1BWF(gedilevel1B, shot_number=shotnum_grass)

#check shot's data structure
head(wf_forest@dt)

```

```{r plot raw waveforms}
#plot raw waveforms
par(mfrow = c(1,1), mar=c(4,4,1,1), cex.axis = 1.5)

plot(wf_forest, relative=FALSE, polygon=TRUE, type="l", lwd=2, col="#FDE725",
    xlab="Waveform Amplitude", ylab="Elevation (m)", main="Raw Waveform: Forest", xlim=c(200,500), ylim=c(200,260))

plot(wf_edge, relative=FALSE, polygon=TRUE, type="l", lwd=2, col="#21908D",
    xlab="Waveform Amplitude", ylab="Elevation (m)", main="Raw Waveform: Edge", xlim=c(200,500), ylim=c(200,260), add = TRUE)

plot(wf_grass, relative=FALSE, polygon=TRUE, type="l", lwd=2, col="#440154",
    xlab="Waveform Amplitude", ylab="Elevation (m)", main="Raw Waveform: Grass", xlim=c(200,500), ylim=c(200,260), add = TRUE)
```

3.2 Plot GEDIL2A Elevation and Height Metrics 
```{r plot height profiles}
#plot waveform with GEDIL2A relative height metrics
plotWFMetrics(level1b = gedilevel1B, level2a = gedilevel2A, shot_number = shotnum_forest, rh=c(25, 50, 75, 90))
plotWFMetrics(gedilevel1B, gedilevel2A, shotnum_edge, rh=c(25, 50, 75, 90))
plotWFMetrics(gedilevel1B, gedilevel2A, shotnum_grass, rh=c(25, 50, 75, 90))
```

Lets use the same elevation range to compare the previous shots
```{r plot height profiles same range}
#plot waveform with GEDIL2A relative height metrics but with a standard scale

plotWFMetrics(gedilevel1B, gedilevel2A, shotnum_forest, rh=c(25, 50, 75, 90), customylim = c(200, 240), main = 'Forest', add = TRUE, colBG = "#FDE725")
plotWFMetrics(gedilevel1B, gedilevel2A, shotnum_edge, rh=c(25, 50, 75, 90), customylim = c(200, 240), main = 'Edge', add = TRUE, colBG = "#21908D")
plotWFMetrics(gedilevel1B, gedilevel2A, shotnum_grass, rh=c(25, 50, 75, 90), customylim = c(200, 240), main = 'grass', add = TRUE, colBG = "#440154")


```



3.3 Plot GEDIL2B Plant Area Index (PAI) and Plant Area Volume Density (PAVD)
```{r get L2B data in R}
#get canopy cover and vertical profile metrics
level2BVPM <- getLevel2BVPM(gedilevel2B)

#check the data structure and 3 first values
head(level2BVPM, 3) 

# check the dimensions
dim(level2BVPM)


#filtering for quality_flag == 1
level2BVPM <- level2BVPM[l2b_quality_flag == 1]
dim(level2BVPM)

#removing NA values 
level2BVPM$pai[level2BVPM$pai==-9999]<-NA # assigning NA to -9999
dim(level2BVPM)
```


```{r plot PAI profiles}
#get pai profile metrics 
level2BPAIProfile <- getLevel2BPAIProfile(gedilevel2B)
dim(level2BPAIProfile)

#filtering for quality_flag == 1
level2BPAIProfile <- level2BPAIProfile[l2b_quality_flag == 1]
dim(level2BPAIProfile)

#convert shot_number from integer to character
level2BPAIProfile$shot_number <- paste0(level2BPAIProfile$shot_number)


#select the previous shots to examine their PAI
smallLevel2BPAIProfile <- level2BPAIProfile[level2BPAIProfile$shot_number %in% sample_shotnums, ]
sample_colors <- c("#FDE725", "#21908D",  "#440154")

#checking the data structure
head(smallLevel2BPAIProfile)[, 1:18]

#plot some shots
PAIplot <- plotPAIshots(smallLevel2BPAIProfile, col_vector = sample_colors, cutUntil0 = TRUE, idcolumn = 'shot_number')
```


```{r plot PAI track}
#plot Level2B PAI Profile 
gPAIprofile <- plotPAIProfile(level2BPAIProfile, beam="BEAM0101", elev=TRUE)
```

```{r plot PAVD profiles}
#get pavd profile metrics 
level2BPAVDProfile <- getLevel2BPAVDProfile(gedilevel2B)
dim(level2BPAVDProfile)

#filtering for quality_flag == 1
level2BPAVDProfile <- level2BPAVDProfile[l2b_quality_flag == 1]
dim(level2BPAVDProfile)

#convert shot_number from integer to character
level2BPAVDProfile$shot_number <- paste0(level2BPAVDProfile$shot_number)

#select the previous shots to examine their PAI
smallLevel2BPVDProfile <- level2BPAVDProfile[as.character(level2BPAVDProfile$shot_number) %in% sample_shotnums, ]
sample_colors <- c("#FDE725", "#21908D",  "#440154")

#checking the data structure
head(smallLevel2BPVDProfile)[, 1:18]

#plot some shots
PAVDplot <- plotPAVDshots(smallLevel2BPVDProfile, col_vector = sample_colors, cutUntil0 = TRUE, idcolumn = 'shot_number')
```


```{r plot PAVD track}
#plot Level2B PAVD Profile
gPAVDprofile<-plotPAVDProfile(level2BPAVDProfile, beam="BEAM0101", elev=TRUE)
```

## SECTION 4: Create Raster Layers 
4.1 Compute Descriptive Statistics 
```{r raster metrics}
#compute a series of statistics for GEDI relative height metrics 
rh100metrics_st <- polyStatsLevel2AM(level2AM_qual, func=SetOfMetrics(rh100))
print(rh100metrics_st) # Check the results

#compute a series of statistics for GEDI canopy cover metrics 
cover_metrics_st <- polyStatsLevel2BVPM(level2BVPM, func=SetOfMetrics(cover))
print(cover_metrics_st) # Check the results

```

4.1 Compute Descriptive Height Stats 
```{r raster summary}
#compute a series of statistics for GEDI relative height metrics 
rh100metrics <- gridStatsLevel2AM(level2AM_qual, func=SetOfMetrics(rh100), res=0.010)
rh100metrics # the results here are rasters
```

```{r plot rasters}
#plot rh100 stats
rh100maps <- levelplot(rh100metrics,
                     layout=c(4, 1),
                     margin=FALSE,
                     xlab = "Longitude (degree)", ylab = "Latitude (degree)",
                     colorkey=list(
                       space='right',
                       labels=list(at=seq(0, 46, 2), font=4),
                       axis.line=list(col='black'),
                       width=1),
                     par.settings=list(
                       strip.border=list(col='gray'),
                       strip.background=list(col='gray'),
                       axis.line=list(col='gray')
                     ),
                     scales=list(draw=TRUE),
                     col.regions=viridis,
                     at=seq(0, 46, len=101),
                     names.attr=c("rh100 min","rh100 max","rh100 mean", "rh100 sd"))
rh100maps
```


```{r compare points raster}
#plot rh100 stats over the map
rh100metrics@crs <- CRS("+proj=longlat +datum=WGS84 +no_defs")
sample_level2AM_qual <- level2AM_qual[sample(size = 100, nrow(level2AM_qual)), ]
rhrastvalues <- unique(as.numeric(rh100metrics@data@values))
rhPal<-  colorNumeric(palette = "viridis", reverse = FALSE,
                      domain = rhrastvalues, na.color = "transparent")


leaflet() %>% addTiles() %>% 
  
  addRasterImage(rh100metrics[['mean']], colors = rhPal, opacity = .7, group = "Mean") %>%
  addRasterImage(rh100metrics[['min']], colors = rhPal, opacity = .7, group = "Min") %>%
  addRasterImage(rh100metrics[['max']], colors = rhPal, opacity = .7, group = "Max") %>%
  addRasterImage(rh100metrics[['sd']], colors = rhPal, opacity = .7, group = "SD") %>%
  
  
  addLegend(pal = rhPal, values = rhrastvalues,
            position = 'topleft',
            title= ""#, opacity = .3
            #, labFormat = labelFormat(transform = function(x) sort(x, decreasing = TRUE))
  ) %>%
     addCircleMarkers(data = sample_level2AM_qual, lng = sample_level2AM_qual$lon_lowestmode, 
                   lat = sample_level2AM_qual$lat_lowestmode,
                   popup = paste0('RH100:', round(sample_level2AM_qual$rh100, 2),'<br>',
                                  'Shot:', sample_level2AM_qual$shot_number), 
                   radius = .2, group = 'GEDI', 
                   color = ~rhPal(sample_level2AM_qual$rh100))  %>%
 
  addLayersControl(position = 'topleft',
                   overlayGroups = c('Mean', 'Min', 'Max', 'SD', 'GEDI'),
                   options = layersControlOptions(collapsed = FALSE),
                   baseGroups = c("OpenStreetMap", "Esri.WorldImagery")) %>%
  addProviderTiles( "Esri.WorldImagery", group = "Esri.WorldImagery" )

```



4.2 Compute Descriptive Canopy Cover Stats 
```{r L2B stats}
#compute a series of statistics for GEDI plant area index
pai_metrics <- gridStatsLevel2BVPM(level2BVPM = level2BVPM, func=SetOfMetrics(pai), res=0.010)
fhd_metrics <- gridStatsLevel2BVPM(level2BVPM = level2BVPM, func=SetOfMetrics(fhd_normal), res=0.010)
cover_metrics <- gridStatsLevel2BVPM(level2BVPM = level2BVPM, func=SetOfMetrics(cover), res=0.010)
```

```{r plot raster L2B stats}
#plot pai stats
pai_maps <- levelplot(pai_metrics,
                    layout=c(4, 1),
                    margin=FALSE,
                    xlab = "Longitude (degree)", ylab = "Latitude (degree)",
                    colorkey=list(
                      space='right',
                      labels=list(at=seq(0, 8, 0.2), font=4),
                      axis.line=list(col='black'),
                      width=1),
                    par.settings=list(
                      strip.border=list(col='gray'),
                      strip.background=list(col='gray'),
                      axis.line=list(col='gray')
                    ),
                    scales=list(draw=TRUE),
                    col.regions=viridis,
                    at=seq(0, 8, len=101),
                    names.attr=c("PAI min","PAI max","PAI mean", "PAI sd"))
pai_maps
```

```{r plot PAI}
#plot pai stats over the map
pai_metrics@crs <- CRS("+proj=longlat +datum=WGS84 +no_defs")
sample_level2BPAI <- level2BPAIProfile[sample(size = 100, nrow(level2BPAIProfile)), ]
pairastvalues <- unique(as.numeric(pai_metrics@data@values))
paiPal<-  colorNumeric(palette = "viridis", 
                      domain = pairastvalues, na.color = "transparent")


leaflet() %>% addTiles() %>% 
  addRasterImage(pai_metrics[['mean']], colors = rhPal, opacity = .7, group = "Mean") %>%
  addRasterImage(pai_metrics[['min']], colors = rhPal, opacity = .7, group = "Min") %>%
  addRasterImage(pai_metrics[['max']], colors = rhPal, opacity = .7, group = "Max") %>%
  addRasterImage(pai_metrics[['sd']], colors = rhPal, opacity = .7, group = "SD") %>%
  
  
  addLegend(pal = paiPal, values = pairastvalues,
            position = 'topleft',
            title= ""#, opacity = .
  ) %>%
     addCircleMarkers(data = sample_level2BPAI, lng = sample_level2BPAI$lon_lowestmode, 
                   lat = sample_level2BPAI$lat_lowestmode,
                   popup = paste0('PAI-5m: ' , round(sample_level2BPAI$pai_z0_5m, 2),'<br>',
                                  'Shot: ', sample_level2BPAI$shot_number), 
                   radius = .2, group = 'GEDI', 
                   color = ~paiPal(sample_level2BPAI$pai_z0_5m))  %>%
 
  addLayersControl(position = 'topleft',
                   overlayGroups = c('Mean', 'Min', 'Max', 'SD', 'GEDI'),
                   options = layersControlOptions(collapsed = FALSE),
                   baseGroups = c("OpenStreetMap", "Esri.WorldImagery")) %>%
  addProviderTiles( "Esri.WorldImagery", group = "Esri.WorldImagery" )

```


```{r FHD maps}
#plot fhd stats
fhd_maps <- levelplot(fhd_metrics,
                    layout=c(2, 2),
                    margin=FALSE,
                    xlab = "Longitude (degree)", ylab = "Latitude (degree)",
                    colorkey=list(
                      space='right',
                      labels=list(at=seq(0, 3.5, 0.25), font=4),
                      axis.line=list(col='black'),
                      width=1),
                    par.settings=list(
                      strip.border=list(col='gray'),
                      strip.background=list(col='gray'),
                      axis.line=list(col='gray')
                    ),
                    scales=list(draw=TRUE),
                    col.regions=viridis,
                    at=seq(0, 3.5, len=101),
                    names.attr=c("FHD min","FHD max","FHD mean", "FHD sd"))
fhd_maps
```


```{r plot cover}
#plot cover stats
cover_maps <- levelplot(cover_metrics,
                    layout=c(2, 2),
                    margin=FALSE,
                    xlab = "Longitude (degree)", ylab = "Latitude (degree)",
                    colorkey=list(
                      space='right',
                      labels=list(at=seq(0, 1, 0.05), font=4),
                      axis.line=list(col='black'),
                      width=1),
                    par.settings=list(
                      strip.border=list(col='gray'),
                      strip.background=list(col='gray'),
                      axis.line=list(col='gray')
                    ),
                    scales=list(draw=TRUE),
                    col.regions=viridis,
                    at=seq(0, 1, len=101),
                    names.attr=c("CCF min","CCF max","CCF mean", "CCF sd"))
cover_maps
```



## SECTION 5: Compare two vegetation covers
5.1 Define polygon areas

We are definig two contrasting polygons, humand grasslands and forests. For this you can read polygons in WKT format (text) or from vecor/raster files using functions like raster::raster("mytif.tif") or rgdal::readOGR('mygpkg.gpkg')

```{r define polygons}
# Provide study areas
wkt_deforested <- "POLYGON((-75.001718%200.23935,-74.998442%200.238726,-74.991579%200.237478,-74.986899%200.236073,-74.981439%200.234513,-74.983311%200.231079,-74.980659%200.22749,-74.975355%200.22234,-74.966463%200.225149,-74.965371%200.229831,-74.969115%200.234513,-74.977695%200.241223,-74.977383%200.244188,-74.969895%200.240287,-74.967243%200.23701,-74.958976%200.234357,-74.958352%200.228114,-74.962564%200.22156,-74.966463%200.220936,-74.967555%200.217971,-74.966775%200.214538,-74.966307%200.213913,-74.96116%200.20892,-74.955076%200.209232,-74.957884%200.204862,-74.96038%200.200961,-74.962252%200.195343,-74.966151%200.196591,-74.967711%200.201117,-74.967087%200.205799,-74.965527%200.209856,-74.970675%200.217035,-74.973764%200.219715,-74.977586%200.221665,-74.982343%200.22346,-74.983045%200.226113,-74.986165%200.229,-74.991157%200.231341,-74.989861%200.230941,-74.99232%200.231292,-74.99388%200.231994,-74.994582%200.234608,-74.995791%200.236481,-74.998794%200.237066,-75.000822%200.237027,-75.001718%200.23935))"

wkt_forest <-  "POLYGON((-74.927009%200.104278,-74.892711%200.149222,-74.830976%200.101157,-74.874627%200.049971,-74.927009%200.104278))"


# Convert WKT into R polygons
forest_poly <- readWKT(gsub('%20', ' ', wkt_forest))
deforested_poly <- readWKT(gsub('%20', ' ', wkt_deforested))

## You can load your own polygons loading shapefiles:
# forest_poly <- rgdal::readOGR('/path/to', 'myforestlayer.shp')
```


```{r filter plot points}
# Filter coordinates
deforested_pts <- level2AM_qual[na.omit(raster::extract(deforested_poly, level2AM_qual[, c('lon_lowestmode', 'lat_lowestmode')]))$point.ID,]
forest_pts <- level2AM_qual[na.omit(raster::extract(forest_poly, level2AM_qual[, c('lon_lowestmode', 'lat_lowestmode')]))$point.ID,]

ptsPal<-  colorNumeric(palette = "viridis", reverse = TRUE,
                      domain = c(deforested_pts$rh100, forest_pts$rh100), na.color = "transparent")


# Mapping the areas and GEDI points
leaflet() %>%
  addCircleMarkers(data = deforested_pts,
                   lng = deforested_pts$lon_lowestmode,
                   lat = deforested_pts$lat_lowestmode,
                   radius = 1, opacity = 0.5,
                   color = ~ptsPal(rh100),
                   popup = paste0("shot_number: ", deforested_pts$shot_number)) %>%
  addCircleMarkers(data = forest_pts,
                   forest_pts$lon_lowestmode,
                   forest_pts$lat_lowestmode,
                   radius = 1, opacity = 0.5, 
                   color = ~ptsPal(rh100),
                   popup = paste0("shot_number: ", forest_pts$shot_number))  %>%
  addPolygons(data = deforested_poly, color = 'red', fill = NA, group = 'Deforested' ) %>% 
  addPolygons(data = forest_poly, color = 'blue', fill = NA, group = 'Forest' ) %>% 
    addLegend(pal = paiPal, values = pairastvalues,
            position = 'topleft', title= "Height (m)") %>%
  addScaleBar(options = list(imperial = FALSE)) %>%
  addProviderTiles(providers$Esri.WorldImagery) %>%
  addLegend(colors = c("blue", "red"), labels = c("Forest", "Grasslands"))

```

Explore how the energy curves looks in both polygons
```{r compare plots}
rh100Pts <- rbind.data.frame(cbind(deforested_pts, label = 'Deforested', clr = "#F8766D"),
                                   cbind(forest_pts, label = 'Forest', clr = '#00BFC4'))

plotRh100 <- plot_rh_profiles(rh100Pts, idcolumn = 'shot_number', col_vector = rh100Pts$clr, bunch = TRUE)
```

Now we will compare the profiles and max. height (rh100) in both polygons
```{r boxplot}
ggp <- ggplot(data = rh100Pts[, c('shot_number', 'rh100', 'label', 'clr')], 
       aes(x = label, y = rh100, color = label)) +
  geom_boxplot(fill=0.8, outlier.size=0.5, color = 'black') +  
  geom_point(position=position_jitter(width=0.3), alpha=0.8) +
  guides(colour = guide_legend(override.aes = list(alpha=1))) + 
  labs(x = "Area", y = "Heigh (m)t", title = 'Max. heigth', color = 'Area')
ggp
```


## SECTION 6: Model applications

Species Distribution Model fake species (created artificial presence/absence points in QGIS)
```{r spp points}
sp_locations <- readOGR(dsn= "data/fake_sp.shp")
head(sp_locations)

plot(sp_locations[sp_locations$Presence == 0,],col='red',pch=16)
plot(sp_locations[sp_locations$Presence == 1,],col='blue',pch=16)
```

Add environmental variable raster brick and then add GEDI rasters to it
```{r}
bioclim_data <- brick(paste0(wd,"/data/bioclim_data_AOI2.tif"))

#change names of layers
names(bioclim_data) <- c("MeanAnnualTemp", "TempAnnualRange", "MeanAnnualPrecip")

#take a look
plot(bioclim_data$MeanAnnualTemp)
plot(bioclim_data$TempAnnualRange)
plot(bioclim_data$MeanAnnualPrecip)
max_canopyht <- rh100metrics$max

#look at extents
bioclim_data
rh100metrics

#resample so that we can combine rasters
#rh100_resample<-resample(rh100metrics$max, bioclim_data, method="ngb")
bioclim_resample<-resample(bioclim_data, max_canopyht, method="ngb")

#add the GEDI max canopy height layer to the Bioclim raster stack
preds <- addLayer(bioclim_resample, max_canopyht)
plot(preds$max)
```


Using the sdm package to run a Linear model and Random Forest model
```{r}
#get data into format for sdm package and add 60 random background points
d<- sdmData(Presence~ ., sp_locations, predictors=preds) 

#what models are available in the sdm package?
getmethodNames()

#Let's try 2: a generalized linear model and a random forest model
m1 <- sdm(Presence~.,data=d,methods=c('glm', 'rf'))
m1
```



```{r}
#evaluate with ROC curves
roc(m1)
```


```{r}
#variable importance Plot
getVarImp(m1)
```


```{r}
#predict
p1 <- predict(m1,newdata=preds) 
plot(p1)
```