---
title: "AGU_GEDI_Workshop"
author: "Melissa Rose, Ivan Gonzalez, Jenna Keanny"
date: '2022-11-21'
output: html_document
---

SECTION 0: PREPARE R ENVIRONMENT
0.1: Libraries
```{r, warning=FALSE, message=FALSE}
#define which libraries to use
library(raster)   
library(devtools) 
library(sf) 
library(sp)
library(data.table)
library(lidR)
library(httr)
library(tidyr)
library(curl)
library(hdf5r)
library(leaflet)
library(leafsync)
```

0.2: Directory and Subdirectories 
```{r}
#define folder paths
wd <- getwd()

gL1B_dir <- './data/GEDI01_B.002/'
gL2A_dir <- './data/GEDI02_A.002/'
gL2B_dir <- './data/GEDI02_B.002/'
shp_dir <- './data/shapefile/'
```

0.3: Source Scripts
```{r, warning=FALSE, message=FALSE}
#source rGEDI functions in scripts directory
source('./scripts/gedi_functions.R')
```


SECTION 1: DOWNLOAD GEDI DATA FOR A REGION OF INTEREST
1.1: Define region of interest and date range 
```{r}
#ROI: Columbia
#specify bounding box coordinates for ROI
ll_lon <- -75  #lower left longitude 
ll_lat <- 0    #lower left latitude 
ur_lon <- -74  #upper right longitude
ur_lat <- 1    #upper right latitude 

#convert into bounding box coordinates
bbox <- paste(ll_lon, ll_lat, ur_lon, ur_lat, sep = ',')

#specify the date range to filter orbits by
daterange=c("2020-05-01","2020-07-31")
```

1.2: Visualize region of interest
```{r}
#view footprint locations for entire orbit on map 
leaflet() %>%
  addRectangles(
    lng1=ll_lon, lat1=ll_lat,
    lng2=ur_lon, lat2=ur_lat,
    color = "red",
    fillColor = "transparent") %>%
  addScaleBar(options = list(imperial = FALSE)) %>%
  addProviderTiles(providers$Esri.WorldImagery) %>%
  addLegend(colors = "red", labels = "Bounding Box") 
```

1.3: Use gedi_finder to find overlapping GEDI orbits 
```{r, warning=FALSE, message=FALSE}
#create a list of all overlapping orbits for each GEDI data product 
gL1B <- gedi_finder('GEDI01_B.002', bbox)  
gL2A <- gedi_finder('GEDI02_A.002', bbox)
gL2B <- gedi_finder('GEDI02_B.002', bbox)
```

1.4 Use gedi_temp_filter to subset orbit list within date range
```{r}
#subset orbit list using daterange
gL1B <- gedi_temp_filter(gL1B, 'GEDI01_B.002', daterange)
gL2A <- gedi_temp_filter(gL2A, 'GEDI02_A.002', daterange)
gL2B <- gedi_temp_filter(gL2B, 'GEDI02_B.002', daterange)
```

1.5: Use gedi_download to download GEDI data (DO NOT RUN)
```{r}
#download data product to their output data directory 
gedi_download(filepath=gL1B,outdir=gL1B_dir)
gedi_download(filepath=gL2A,outdir=gL2A_dir)
gedi_download(filepath=gL2B,outdir=gL2B_dir)
```


SECTION 2: CLIP GEDI DATA PRODUCTS TO REGION OF INTEREST 
2.1 Read GEDI data
```{r}
#read the level 1B, 2A, and 2B for a single orbit
gedilevel1B <- readLevel1B(level1Bpath = paste0(gL1B_dir,"GEDI01_B_2020121050305_O07826_02_T03778_02_005_01_V002.h5"))
gedilevel2A <- readLevel2A(level2Apath = paste0(gL2A_dir,"GEDI02_A_2020121050305_O07826_02_T03778_02_003_01_V002.h5"))
gedilevel2B <- readLevel2B(level2Bpath = paste0(gL2B_dir,"GEDI02_B_2020121050305_O07826_02_T03778_02_003_01_V002.h5"))
```

2.2 Clip data within bbox coordinates (DO NOT RUN)
```{r}
gedilevel1B_clip <- clipLevel1B(gedilevel1B, ll_lon, ur_lon, ll_lat, ur_lat,
                                output = paste0(gL1B_dir,"GEDI01_B_2020121050305_O07826_02_T03778_02_005_01_V002_clip.h5"))
gedilevel2A_clip <- clipLevel2A(gedilevel2A, ll_lon, ur_lon, ll_lat, ur_lat, 
                                output = paste0(gL2A_dir,"GEDI02_A_2020121050305_O07826_02_T03778_02_003_01_V002_clip.h5"))
gedilevel2B_clip <- clipLevel2B(gedilevel2B, ll_lon, ur_lon, ll_lat, ur_lat, 
                                output = paste0(gL2B_dir,"GEDI02_B_2020121050305_O07826_02_T03778_02_003_01_V002_clip.h5"))
```

```{r}
gedilevel1B_clip <- readLevel1B(level1Bpath = paste0(gL1B_dir,"GEDI01_B_2020121050305_O07826_02_T03778_02_005_01_V002_clip.h5"))
gedilevel2A_clip <- readLevel2A(level2Apath = paste0(gL2A_dir,"GEDI02_A_2020121050305_O07826_02_T03778_02_003_01_V002_clip.h5"))
gedilevel2B_clip <- readLevel2B(level2Bpath = paste0(gL2B_dir,"GEDI02_B_2020121050305_O07826_02_T03778_02_003_01_V002_clip.h5"))
```

2.3 Retrieve coordinates for each GEDI footprint in clipped dataset
```{r}
#use GEDIL1B data to find coordinates for each shot_number 
level1bGeo <- getLevel1BGeo(level1b=gedilevel1B_clip,select=c("elevation_bin0"))
head(level1bGeo)

#convert shot_number from integer to character
level1bGeo$shot_number <- paste0(level1bGeo$shot_number)

#remove NA coordinate values
level1bGeo <- na.omit(level1bGeo)

#convert level1bGeo to spatial object
level1bGeo_spdf <- SpatialPointsDataFrame(cbind(level1bGeo$longitude_bin0, level1bGeo$latitude_bin0),
                                        data=level1bGeo)
#export orbit footprint locations as shapefile
#shapefile(level1bGeo_spdf, paste0(shp_dir, "GEDI01_B_2020121050305_O07826_02_T03778_02_005_01_V002_clip"))
```

2.4 Visualize location of clipped GEDI footprints  
```{r}
#view footprint locations for entire orbit on map 
leaflet() %>%
  addRectangles(
    lng1=ll_lon, lat1=ll_lat,
    lng2=ur_lon, lat2=ur_lat,
    color = "red",
    fillColor = "transparent") %>%
  addCircleMarkers(level1bGeo$longitude_bin0,
                   level1bGeo$latitude_bin0,
                   radius = 1,
                   opacity = 0.5,
                   color = "blue", 
                   popup = paste0("shot_number: ", level1bGeo$shot_number))  %>%
  addScaleBar(options = list(imperial = FALSE)) %>%
  addProviderTiles(providers$Esri.WorldImagery) %>%
  addLegend(colors = c("red", "blue"), labels = c("Bounding Box", "GEDI Footprints")) 
```

```{r}
sample_region <- level1bGeo[shot_number %in% c("78260000200055294", "78260000200055295", "78260000200055296", "78260000200055297", "78260000200055298")]

#view footprint in sample region 
leaflet() %>%
  addRectangles(
    lng1=-74.079, lat1=0.3220,
    lng2=-74.0765, lat2=0.32425,
    color = "grey",
    fillColor = "transparent") %>%
  addCircleMarkers(sample_region$longitude_bin0,
                   sample_region$latitude_bin0,
                   radius = 25,
                   weight = 2,
                   fillColor = "transparent",
                   color = "blue", 
                   popup = paste0("shot_number: ", sample_region$shot_number))  %>%
  addScaleBar(options = list(imperial = FALSE)) %>%
  addProviderTiles(providers$Esri.WorldImagery) %>%
  addLegend(colors = c("grey", "blue"), labels = c("Sample Region", "GEDI Footprints")) 
```

SECTION 3: Visualize GEDI Data Products 
3.1 Plot GEDIL1B Full Waveform 
```{r}
#select shot_numbers for disturbed, edge, and intact forest
shot_number <- "78261100200053382"

wf <- getLevel1BWF(level1b, shot_number=shot_number)

par(mfrow = c(1,2), cex.axis = 1.5)
plot(wf, relative=FALSE, polygon=TRUE, type="l", lwd=2, col="forestgreen",
xlab="", ylab="Elevation (m)")


#plotWFMetrics(gedilevel1B_clip, gedilevel2A_clip, shot_number, rh=c(25, 50, 75, 90))

```

```{r}
level1b=gedilevel1B_clip
level2a=gedilevel2A_clip
shot_number="78261100200053382"
rh=c(25, 50, 75)

  # Avoid NOTEs from checking
  elevation = NULL
  oldpar <- par(no.readonly = TRUE)
  on.exit(par(oldpar)) 

  # Extracting GEDI full waveform for a giving shotnumber
  wf <- getLevel1BWF(level1b, shot_number=shot_number)

  level2AM<-getLevel2AM(level2a)
  shotid_mask = which(level2AM$shot_number == shot_number)

  level2_shot = level2AM[shotid_mask,]
  ground_z = level2_shot$elev_lowestmode
  rhs = paste0(as.character(c("rh0", as.character(paste0("rh",rh)), "rh100")))
  rh = level2_shot[, rhs, with=FALSE]
  rh_z = rh + ground_z


  top_z = level2AM[shotid_mask,]$elev_highestreturn

  range_energy = range(wf@dt$rxwaveform)
  range_abs_diff = abs(diff(range_energy))

  requireNamespace("data.table")

  range_z = range(rh_z)
  min_z = min(range_z)
  max_z = max(range_z)
  diff_z = abs(diff(range_z))
  wf_between = wf@dt[elevation %between% range_z,,]
  energy_offset = min(range_energy)
  energy_no_offset = (wf_between$rxwaveform - energy_offset)
  cumsum_energy = cumsum(rev(energy_no_offset))

  range_cumsum = range(cumsum_energy)
  range_abs_diff_cumsum = abs(diff(range_cumsum))
  energy_cum_normalized = ((cumsum_energy)/(range_abs_diff_cumsum/range_abs_diff))+energy_offset

  par(mar = c(5, 4, 4, 4) + 0.3)
  offset = diff_z*0.2
  ymin = min_z-offset
  ymax = max_z+offset
  wf_interest=wf@dt[wf@dt$elevation >= ymin & wf@dt$elevation <= ymax,]$rxwaveform
  qts=quantile(wf_interest, c(0.05, 1), type=1)


  z_masked = rev(wf_between$elevation)


  ticks = seq(min_z, max_z, length=4)
  ticks_label = format(ticks-min_z, digits = 2)

  rh_closest_en = list()
  for (i in 1:length(rh_z)) {
    absdiff_rh = abs(z_masked-rh_z[[i]])
    rh_closest_en[[names(rh_z)[[i]]]] = which(absdiff_rh==min(abs(absdiff_rh)))
  }

  # Make marks for RH based in a point
  mark = function(x, y, ...) {
    arrows(x, y, x, min_z, length=.1, code = 3)
  }

  # Find mid y for rh labels
  ymidpoint = function(x) {
    x-(x-min_z)/2
  }


  plot(wf, relative=FALSE, polygon=TRUE, type="l", lwd=2, col="forestgreen",
       xlab="Waveform Amplitude", ylab="Elevation (m)", ylim=c(ymin, ymax), xlim=qts+c(0, 0.1*abs(diff(qts))), ...)
  par(new=TRUE)
  plot(energy_cum_normalized, z_masked, lwd=2, axes=F, bty="n", type="l", xlab = "", ylab = "", ylim=c(ymin, ymax), xlim=qts)
  axis(side=4, at = ticks, labels=ticks_label)
  mtext("Height (m)", side=4, line=2)
  for (i in 2:(length(rh_z)-1)) {
    mark(energy_cum_normalized[rh_closest_en[[i]]], rh_z[[i]])
    text(energy_cum_normalized[rh_closest_en[[i]]], ymidpoint(rh_z[[i]]), toupper(names(rh_z)[[i]]), pos = 2)
  }
  text(qts[2]-diff(qts)/2, rh_z[[length(rh_z)]], "RH100", pos=3)
  abline(rh_z[[length(rh_z)]], 0, lty="dashed")
  text(qts[2]-diff(qts)/2, rh_z[[1]], "RH0", pos=1)
  abline(rh_z[[1]], 0, lty="dashed")
```



SECTION 4: Create Raster



